---
output:
  pdf_document:
    includes:
      in_header: "review-packet-02-final-preamble.tex"
fontfamily: cochineal
fontsize: 12pt
header-includes:
   - \linespread{1.05}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tufte)
library(kableExtra)
library(tidyverse)
```

**Problem 7**. Compute the F statistic for a one-way analysis of variance investigating whether differences exist between the three levels of the factor below and calculate the effect size associated with this test.

```{r p7, echo=FALSE}


p7 <- data.frame(Group1 = c(8,7,6,9,8,7,8,3),
                 Group2 = c(7,6,5,6,4,5,6,4),
                 Group3 = c(2,5,4,4,6,5,3,2))

p7 %>%
  kbl() %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

p7_long <- p7 %>% 
  pivot_longer(Group1:Group3, names_to = "group", values_to = "dv")

summary(aov(dv ~ group, data = p7_long))

```

\newpage

**Problem 8**. Compute the F statistic for a one-way analysis of variance investigating whether differences exist between the three levels of the factor below and calculate the effect size associated with this test.

```{r p8, echo=FALSE}
p8 <- data.frame(group1 = c(55,64,34,53,67,86,67,67,65,48,91),
                 group2 = c(45,34,43,53,65,76,34,36,34,88,87),
                 group3 = c(92,34,45,54,65,76,35,98,76,45,51))

p8 %>%
  kbl() %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

p8_long <- p8 %>% 
  pivot_longer(group1:group3, names_to = "group", values_to = "dv")

summary(aov(dv ~ group, data = p8_long))

```

\newpage

**Problem 9**. Compute the F statistic for a one-way analysis of variance investigating whether differences exist between the four levels of the factor below and calculate the effect size associated with this test

```{r p9, echo=FALSE}
p9 <- data.frame(Group1 = c(2,3,4,3,2,1),
                 Group2 = c(2,0,-1,2,-3,-4),
                 Group3 = c(3,4,1,-1,-2,-3),
                 Group4 = c(1,0,0,3,2,6))

p9 %>%
  kbl() %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

p9_long <- p9 %>% 
  pivot_longer(Group1:Group4, names_to = "group", values_to = "dv")

summary(aov(dv ~ group, data = p9_long))

```

\newpage

**Problem 10**. Compute the F statistic for a one-way analysis of variance investigating whether differences exist between the three levels of the factor below and calculate the effect size associated with this test.


```{r p10, echo=FALSE}

p10 <- data.frame(Group1 = c(6,7,6,5,6,6,6, NA, NA, NA, NA),
                  Group2 = c(5,4,3,7,8,5,6,4,3,7,6),
                  Group3 = c(9,4,5,2,3,4, NA, NA, NA, NA, NA))

p10 %>%
  kbl() %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

p10_long <- p10 %>% 
  pivot_longer(Group1:Group3, names_to = "group", values_to = "dv")

summary(aov(dv ~ group, data = p10_long))

```
