---
output:
  pdf_document:
    includes:
      in_header: "review-packet-01-midterm-preamble.tex"
fontfamily: cochineal
fontsize: 12pt
header-includes:
   - \linespread{1.05}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tufte)
library(kableExtra)
library(tidyverse)
```

# Midterm Review Practice Packet

This practice packet covers all the main computations needed to master for the midterm exam. I suggest printing this packet and practicing the calculations by-hand with a calculator and the _Midterm Forumla Sheet_, as you would during the Midterm Exam. Solutions are presented on the last two pages.

\noindent\rule{\textwidth}{0.4pt}

**Problem 1**. Calculate the mean, median, and mode of the data set below.  

$$1, 3, 7, 6, 3, 5, 1, 7, 8, 7$$
\newpage

**Problem 2**. Calculate the mean, median, and mode of the data set below.

$$62, 51, 76, 92, 51, 16, 22, 51, 23, 11$$

\newpage

**Problem 3**. Calculate the range, standard deviation, and variance of the data set in _Problem 1_.

$$1, 3, 7, 6, 3, 5, 1, 7, 8, 7$$

\newpage

**Problem 4**. Calculate the range, standard deviation, and variance of the data set in _Problem 2_.

$$62, 51, 76, 92, 51, 16, 22, 51, 23, 11$$

\newpage

**Problem 5**. Calculate the correlation between the two variables from _Problem 1_ and _Problem 2_.

```{r echo=FALSE}

corr_data <- data.frame(x = c(1, 3, 7, 6, 3, 5, 1, 7, 8, 7),
                        y = c(62, 51, 76, 92, 51, 16, 22, 51, 23, 11))

kbl(corr_data) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

```

\newpage

**Problem 6**. Calculate the correlation between the two variables in the data set below.


```{r echo=FALSE}

corr_data2 <- data.frame(x = c(10, 1, 5, 4, 8, 11, 2, -3, -5, -2),
                         y = c(-2, 5, 1, -2, -6, 7, -2, 4, 6, 2))

kbl(corr_data2) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

```

\newpage

**Problem 7**. Calculate the internal consistency of the scale below.

```{r ic-1, echo=FALSE}

ic_dat1 <- data.frame(item1 = c(6, 11, 12, 5, 3, 6, 7, 3, 10, 9),
                      item2 = c(7, 10, 11, 4, 4, 4, 5, 4, 8, 3))

kbl(ic_dat1) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

```

\newpage

Use this page as extra space for your calculations for _Problem 7_.

\newpage

**Problem 8**. Calculate the internal consistency of the scale below.

```{r ic-2, echo=FALSE}

ic_dat2 <- data.frame(item1 = c(1, 5, 4, 3, 1, 5, 6, 4, 3, 1),
                      item2 = c(1, 4, 5, 4, 2, 9, 1, 4, 2, 2),
                      item3 = c(2, 1, 4, 4, 4, 4, 6, 4, 3, 1))

kbl(ic_dat2) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")

```

\newpage

Use this page as extra space for your calculations for _Problem 8_.

\newpage

**Problem 9**. Intelligence (as measured by IQ) in the population is normally distributed with a mean of $\mu = 100$ and a standard deviation of $\sigma = 15$. If you take the IQ test and get a score of 118. What is your _z_-score, and how can you interpret this value?

\newpage

**Problem 10**. Your friend takes the same IQ test that you took in _Problem 9_. He is ashamed to tell you his IQ (i.e., his raw score), but after some pressure he concedes the _z_-score associated with his raw score: -2.8. Calculate your friend’s raw IQ score.

\newpage

# Solutions

**Problem 1**

- Mean: 4.8

- Median: 5.5

- Mode: 7

**Problem 2**

- Mean: 45.5

- Median: 51

- Mode: 51

**Problem 3**

- Range: 7

- Standard deviation: 2.6

- Variance: 6.8

**Problem 4**

- Range: 81

- Standard deviation: 27.0

- Variance: 730.6

**Problem 5**

- Correlation: -0.02

**Problem 6**

- Correlation: -0.38

**Problem 7**

- Cronbach’s alpha: 0.86

**Problem 8**

- Cronbach’s alpha: 0.58

\newpage

# Solutions

**Problem 9**

- Z-score: 1.2

- Interpretation: You are 1.2 standard deviations above the mean of intelligence. Because IQ is normally distributed and we know probabilities associated with outcomes under normal curves, only about 16% of the population is smarter than you (approximately).

**Problem 10**

- Raw score: 58



