---
output:
  pdf_document:
    includes:
      in_header: "guide-06-standardization-and-z-scores-preamble.tex"
fontfamily: cochineal
fontsize: 12pt
header-includes:
   - \linespread{1.05}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tufte)
library(kableExtra)
library(tidyverse)
library(bayestestR)
```

# Standardization and _z_-scores

In this _Guide_, we will learn how to calculate _z_-scores. Let’s go ahead and generate some data for us to practice _z_-scoring with. These data are anxiety levels on a 1 (_“not anxious at all”_) to 14 (_“extremely anxious”_) scale of 15 office employees.

```{r anxiety-data, echo=FALSE}
anxiety_scores <- data.frame(
  Name = c(
    "Jim",
    "Pam",
    "Kelly",
    "Ryan",
    "Michael",
    "Holly",
    "Kevin",
    "Dwight",
    "Angela",
    "Toby",
    "Oscar",
    "Andy",
    "Stanley",
    "Phyllis"
  ),
  Anxiety = c(1,
              7,
              12,
              8,
              9,
              5,
              4,
              9,
              7,
              4,
              6,
              13,
              2,
              6)) %>% 
  as_tibble()

knitr::kable(anxiety_scores, "pipe")

```

\begin{center}

z-score formula
$$
z_{i} = \frac{x_{i} - \mu}{\sigma}
$$
\end{center}

Let’s say that we want to know how Dwight’s anxiety level compares to that of his colleagues. We can answer this question by converting his raw score ($9$) into a _z_-score using the formula above. Specifically, we’ll need to subtract Dwight’s raw score from the mean anxiety level of the entire group (the “population” mean, $\mu$, since this group is our entire population of interest) and then divide by the standard deviation of this group ($\sigma$).

Calculating the mean is easy, as we just need to add up all of the anxiety values and divide by $N$. If you do this, you’ll end up with a mean anxiety level of $6.64$.

Next, we’ll need to calculate the standard deviation of anxiety scores using the population formula for standard deviation that we’ve seen in this class previously. As a reminder, here it is:

\begin{center}
Population Standard Deviation
$$
\sigma = \sqrt \frac{\sum(X_{i} - \mu)^2}{N}
$$
\end{center}

I’ll show my work for calculating the standard deviation of this population on the last page, if you’re interested. For now, let me just tell you that $\sigma = 153.22$.

Now, this becomes an easy problem to solve. We can simply plug in Dwight’s anxiety score ($9$, which we represent by $x$), the average level of anxiety in the group ($6.64$, represented by $\mu$), and the standard deviation of anxiety in the group ($3.31$, represented by $\sigma$). Let’s do that now:  

\begin{center}
$$
z_{i} = \frac{x_{i} - \mu}{\sigma} = \frac{9.00-6.64}{3.31} = 0.71
$$
\end{center}

A _z_-score of $0.71$ can be interpreted as follows: Dwight is $0.71$ standard deviations above the mean of anxiety, meaning he is slightly more anxious than average.

Note that, if you were to calculate _z_-scores for every single person in this group, you’ll have an entirely new column of data containing only _z_-scores. We call this _“standardizing”_ data, because the new, “standardized” scores will always have a mean of $0$ and a standard deviation of $1$. This is useful because then each person’s score is a _z_-score that shows how they compare to their peers. This makes it easy to quickly tell who is scoring very high compared to their peers (large, positive _z_-scores) and who is scoring very low compared to their peers (negative _z_-scores).

```{r echo=FALSE, warning=FALSE, fig.width=4, fig.height=3}
library(bayestestR)  # Load it
# Generate a perfect sample
x <- rnorm_perfect(n = 100, mean = 0, sd = 1)

# Visualise it
library(tidyverse)

x %>% 
  density() %>%  # Compute density function
  as.data.frame() %>% 
  ggplot(aes(x=x, y=y)) +
  geom_line() +
  geom_vline(xintercept = 0, ) +
  geom_vline(xintercept = 0.71, linetype="dashed") +
  labs(caption = "Dashed line is Dwight's z-score for anxiety",
       x = "z-score",
       y = '') +
  theme_classic()
```


\newpage

# Calculating Population Standard Deviation

Let’s calculate population standard deviation of this group of office employees. As always, we’ll need to start by finding the mean, which we found was $6.64$. Then, we calculate the deviations of each score from that mean, and then squaring those deviations from the mean. Let’s do that now:

```{r anxiety-scores-z-scored, echo=FALSE}

anxiety_scores_z <- anxiety_scores %>% 
  mutate(deviations = Anxiety - mean(Anxiety),
         squares = deviations^2)

knitr::kable(anxiety_scores_z, "pipe", digits = 3)

```

Now, we need to calculate our _Sum of Squares_ ($\sum(x_{i} - \overline{x})^2$) by adding up all of the values in the _“squares”_ column, which gives us a value of $153.22$.

Finally, to calculate population standard deviation, we need to divide the Sum of Squares by $N$ and then square root that value. Let’s do that now:

\begin{center}
$$
\sigma = \sqrt \frac{\sum(X_{i} - \mu)^2}{N} = \sqrt{\frac{153.22}{14}} = 3.31
$$
\end{center}

So, our population standard deviation ($\sigma$) is $3.31$.


